---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo=FALSE, messages=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(knitr)
library(dplyr)
library(tidyr)
library(lychee)

elec94 <- filter(greens3, year==1994) %>% 
	select(-year,-city_clean)

elec09 <- filter(greens3, year==2009) %>% 
	select(-year,-city_clean)
```


# lychee

The package [`lychee`](https://sumtxt.github.io/lychee/) helps to link and join data frames with key variables that are are similiar but not identical (e.g., a variable with geographic names spelled slightly different or nearby geographic coordinates). Different from the `fuzzyjoin` package, the package [`lychee`](https://sumtxt.github.io/lychee/) does not output all matches given some definition of sufficient similarity, but constructs optimal one-to-one matches minimizing the total difference across all matches. 

The function `linkr()` stacks two data frames and finds an optimal one-to-one pairing of rows in one data frame with rows in the other data frame. The output is a data frame with as many rows as there are in the two datasets and a common identifier for the pairs. The complementary function is `joinr()` which, instead of stacking and assigning a common identifier, joins two data frames similar to the `base::merge()` or `dplyr::full_join()` function. 


### Installation

```{r, eval = FALSE}
# Install development version from GitHub
remotes::install_github("sumtxt/lychee")
```
For more details and to learn how to use this package: [Getting Started with lychee](https://sumtxt.github.io/lychee/articles/lychee.html).


### Usage 

The example below shows how `joinr` finds optimal matches in two data frames (`elec94` and `elec09`) within two groups. The first data frame (`elec94`) lists the strongholds of Germany's green party in the 1994 Federal election (`election=BTW`) and the 1994 European election (`election=EP`). The second data frame lists such strongholds for the 2009 elections. `joinr` merges the two data frames correctly even though the city names are spelled slightly different in the two data frames preventing to merge them via `base::merge()` or `dplyr::full_join()`.

```{r}
elec94

elec09

joinr(elec94,elec09,
	strata="election",
  by="city", 
  suffix=c("94","09"),
  add_distance=TRUE, 
  caliper=12,
  method='lcs',
  full=TRUE)
```








